### 2024-08-15 parts/appendix.tex
enormous rewrite

%! TeX program = xelatex
%! lang = en-US
\section{Proofs of section \ref{preliminaries:sec}}

\begin{proof}[Proof of \cref{decidable-commutative-poly:lemma}, in the polyregular case]
    Let $\Sigma$ be a finite alphabet, endowed with a
    total ordering over its letters,
    and let $f \colon \Sigma^* \to \Rel$ be a
    \kl{$\Rel$-polyregular function}.
    The map $\polysort \colon \Sigma^* \to \Sigma^*$
    that sorts the letters in the input word $w$ according
    to the chosen ordering is a \kl{polyregular function}
    \cite[Proposition II.12]{CDTL23}.
    As a consequence,
    $(f \circ \polysort) \in \ZPoly$.
    Finally, $f$ is \kl{commutative} if and only
    if $f = f \circ \polysort$, which is decidable
    in the case of $\ZPoly$
    \cite[Corollary II.24]{CDTL23}, and more generally in the case 
    of $\ZRat$ \cite[Corollary 3.6 p 38]{BERE10}.
\end{proof}

\begin{proof}[Proof of \cref{decidable-commutative-rat:lemma} in the general case]
    Because $f$ is a \kl{$\Rel$-rational series},
    there exists a $n \in \Nat$,
    vectors $v_1$ and $v_2$,
    and $n$ by $n$ integer matrices $\seqof{M_a}{a \in \Sigma}$,
    such that
    \begin{equation*}
        \forall w \in \Sigma^*,
        f(w) = {}^t v_1 \left(\prod_{i = 1}^{|w|} M_{w_i}\right) v_2 \quad .
    \end{equation*}
    To simplify notations, 
    let us write $M_w \defined \prod_{i = 1}^{|w|} M_{w_i}$,
    and 
    $S \defined \setof{M_w}{w \in \Sigma^+}$ be the
    semigroup generated by $\seqof{M_a}{a \in \Sigma}$.

    The function $f$ is \kl{commutative} if and only if for all $u,v \in
    \Sigma^+$, ${}^t v_1 (M_u M_v  - M_v M_u) v_2 = 0$. The above is a
    polynomial equation involving the coefficients of $M_u$ and $M_v$.
    Let us compute $\overline{S \times S}$ the Zariski closure of $S \times S$.
    The above polynomial vanishes over $S \times S$ if and only
    if it belongs to the Zariski closure, which is decidable
    \cite[\textbf{WHERE}]{HOPW18}.
\end{proof}

@@

\textbf{Redo the proof}.
\begin{proofof}[derivation-stabilises-correct:lem]
    If $P$ is constant, then $P_2 = 0$, and $\Diff{K}{P_1} = 0$,
    hence for $K = 0$, $Q \in \CorrectPoly$.

    Otherwise, let $\alpha$ be the maximal absolute value 
    of coefficients appearing in $P$,
    and $D$ be the maximal number of divisors for
    a monomial in $P$.
    We define $K \defined \alpha \times D + 1$.

    Let $\nu \colon \vec{X} \topartial \Nat$, we are going to prove that
    $\restr{Q}{\nu}$ has \kl{non-negative} \kl{maximal monomials}. To that end,
    let us consider $T$ such a maximal monomial, and let us write $T = \beta
    T_1$ so that $T_1$ has coefficient $1$ and $\beta \in \Rel$. Because we
    will need to normalize other monomials in this way, we define for the rest
    of the proof the function $\mathsf{unit}$ that maps a monomial to the
    monomial obtained by replacing its constant term by $1$ if the monomial is
    non-zero, and leaving the monomial unchanged otherwise.

    Let us write $\mathcal{D}_1$ to be the set of \kl{monomials} of
    $\Diff{K}{P_1}$ that are \kl{divided} by $T$. Similarly, let us write
    $\mathcal{D}_2$ for the set of monomials of $\translate{K}(P_2)$ that are
    \kl{divided} by $T$. Thanks to \cref{derivation-monomials:fact}, we know
    that every element of $\mathcal{D}_2$ divides some element in
    $\mathcal{D}_1$. We write $\mathcal{D}_2^M$ for the set of monomials $S \in
    \mathcal{D}_2$ such that $S$ \kl{divides} $M$.

    Now, remark that the coefficient $\beta$ is obtained as follows:
    \begin{align*}
        \beta &= \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
               + \sum_{M \in \mathcal{D}_2} \restr{M / T_1}{\nu} \\
              &\geq \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
              - \sum_{M \in \mathcal{D}_2} \left|\restr{M / T_1}{\nu}\right|
              \\
              &\geq
              \sum_{M \in \mathcal{D}_1}
              \left[
              \restr{M / T_1}{\nu}
              -
              \sum_{S \in \mathcal{D}_2^M}
              \left|\restr{S/T_1}{\nu}\right|
              \right]
              \\
              &\geq
              \sum_{M \in \mathcal{D}_1}
              \restr{\mathsf{unit}(M)/T_1}{\nu}
              \left[
              K
              - \alpha D
              \right]
              \\
              &\geq 0 \quad . \qedhere
    \end{align*}
\end{proofof}

@@

    We will first prove that \cref{d-t-transl:item} $\implies$
    \cref{d-t-correct:item}, for any choice of $K \in \Nat$. Indeed, let us
    consider some partial valuation $\nu \colon \vec{X} \topartial \Nat$. By
    assumption, $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$, and in
    particular conclude that \kl{maximal monomials} of
    $\translate{K}(\restr{P}{\nu})$ are exactly the \kl{maximal monomials} of
    $\restr{P}{\nu}$ thanks to \cref{translation-maximal:fact}, and are
    therefore \kl{non-negative}.

    Conversely, the implication \cref{d-t-correct:item} $\implies$
    \cref{d-t-transl:item} is proven by induction on $\MaximalMonomials(P)$. If
    $P$ is a constant polynomial, then $P = n$ for some $n \in \Nat$, and we
    conclude that $\translate{0}(P) \in \Nat[\vec{X}]$. Otherwise, $P = P_1 +
    P_2$ where $P_1$ is the sum of the \kl{maximal monomials} of $P$. Using
    \cref{derivation-stabilises-correct:lem}, there exists a computable $K \in
    \Nat$ such that $Q \defined \Diff{K}{P_1} + \translate{K}(P_2)$ belongs to
    $\CorrectPoly$. Since, $\MaximalMonomials(Q) \hoarele
    \MaximalMonomials(P)$, we conclude by induction hypothesis that there
    exists $L \in \Nat$ such that for every partial valuation $\nu \colon
    \vec{X} \topartial \Nat$, $\translate{L}(\restr{Q}{\nu}) \in
    \Nat[\vec{X}]$. Similarly, for all partial functions $\nu \colon \vec{X}
    \topartial \set{0, \dots, K}$ fixing at least one indeterminate of $P$,
    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials(P)$, hence
    there exists $L_\nu \in \Nat$ such that $\translate{L_\nu}(\restr{P}{\nu})
    \in \Nat[\vec{X}]$. Let us define $L_m$ to be the maximum of $L$, and
    $L_\nu$ where $\nu$ ranges over partial functions described above.

    Let us prove that for all partial functions $\nu \colon \vec{X} \topartial
    \Nat$, the translation $\translate{L_m + K}(\restr{P}{\nu})$ belongs to
    $\Nat[\vec{X}]$. Without loss of generality, $\nu$ fixes only
    indeterminates that appear in $P$. If $\nu$ fixes at least one
    indeterminate $X_i$ to a value $\nu(X_i) = k \leq K$, then we can write
    $\nu = \mu [X_i = k]$, where $\mu$ does not fix the value of $X_i$. By
    construction, we know that $\translate{L_i}(\restr{\restr{P}{X_i =
    k}}{\mu})$ belongs to $\Nat[\vec{X}]$, and therefore that $\translate{K +
    L_m}(\restr{P}{\nu})$ does so too. The only case left is when all the
    variables appearing in $\nu$ are greater than $K$.
    In that case, $\nu = \translate{K}(\mu)$ for some partial 
    valuation $\mu$.
    In this case,
    $\translate{K}(\restr{P}{\translate{K}(\mu)}
    = \restr{\translate{K}(P)}{\mu}$
    thanks to 
    \cref{discrete-deriv-linear:fact}.
    Remark that by definition,
    $\translate{K}(P) = P_1 + Q$,
    hence that
    $\translate{K + L_m}(P) = \translate{L_m}(P_1) + \translate{L_m}(Q)$
    belongs to $\Nat[\vec{X}]$.
    Finally,
    we conclude that $\restr{\translate{K+L_m}(P)}{\mu}$
    belongs to $\Nat[\vec{X}]$
    because
    the latter is closed under partial applications.

@@

    Using this $\omega$,
    for any choice of
    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
    there exists an \kl{integer binomial polynomial}
    $P_r \in \Rel[X_1, \dots, X_p]$,
    obtained as
    the difference of the two \kl{natural binomial polynomial} $P_r^+$ and $P_r^-$
    associated to $f_+$ and $f_-$,
    such that $X_1, \dots, X_p \geq 1$:
    \begin{equation*}
        f\left(
        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
        = P_r(X_1, \dots, X_p) \quad .
    \end{equation*}
    We conclude by selecting different values of $r$ and
    setting some subsets of variables to be assigned to the constant $0$.
\end{proofof}

@@

\begin{proofof}[z-poly-commutative-encoding:remark]
    We are going to prove that if $f \in \ZPoly[k]$ is polyblind, 
    then the \cref{z-poly-commutative-encoding:conj} holds when allowing 
    for non polyregular functions $\mathsf{enc}$ and $\mathsf{dec}$.

    Let us consider $f \in \ZPoly[k]$ that is polyblind.
    Using \cite[Theorem 6.12]{DOUE23},
    we know that there exists a family of $\MSO$ formulas
    $\seqof{\psi_{i,j}(x)}{1 \leq i,j \leq n}$,
    and coefficients $\seqof{z_i}{1 \leq i \leq n}$,
    such that:
    \begin{equation*}
        f = \sum_{i = 1}^n z_i \times \prod_{j = 1}^n \vcount{\psi_{ij}(x)}
        \quad .
    \end{equation*}

    Let us now consider a monoid $M$ that recognizes all formulas $\psi_{ij}$
    for $1 \leq i,j \leq n$ with a morphism $\mu \colon \Sigma^* \to M$. Let us
    consider the map $\mathsf{t} \colon \Sigma^* \to (M^3)^*$, that associates
    to a word $w \in \Sigma^*$ the word $\mathsf{t}(w) \defined \prod_{i =
    1}^{|w|} ( \mu(w_{< i}), \mu(w_i), \mu(w_{> i}))$. That is, to every letter
    one associates its left and right context in the monoid.

    Let us define $\mathsf{s} \colon \Sigma^* \to (\Nat)^{M^3}$
    by letting $\mathsf{s}(w)$ be the Parikh image of $\mathsf{t}$, that is,
    counting the number of occurrences of every letter.
    It is clear how to define the polyregular function $g \colon (\Nat)^{M^3} \to \Rel$
    that makes the following diagram commute, at is it even a polynomial function
    with $M^3$ indeterminates:
    \begin{center}
        \begin{tikzcd}
            (M^3)^* 
            \arrow{dr}[below,anchor=center,rotate=-39,yshift=-3pt]{\mathsf{count}}
            & \Sigma^* \arrow{l}[above]{\mathsf{t}} \arrow{r}{f} \arrow{d}[left]{\mathsf{s}} & \Rel \\
            & \Nat^{M^3} \arrow{ur}[below]{g} & \\
        \end{tikzcd}
    \end{center}
    One problem is that the map $\mathsf{s}$ is not surjective, because of
    potential dependencies between the types encountered in $w$.
    Let us however notice that one can build a context-free grammar
    recognizing exactly the image of $\mathsf{t}$. Namely,
    \begin{align*}
        S &\to S^{1,m,1} &  m \in M \\
        S^{m_r, m, m_l} &\to (m_l, m, m_r) & \exists a \in \Sigma, \mu(a) = m \\
        S^{m_r, m, m_l} &\to 
        S^{m_r, m_x, m_y m_l}
        S^{m_r m_x, m_y, m_l}
                        & m_x m_y = m
    \end{align*}

    Using Parikh's Theorem \cite{PARI66},
    we conclude that the image of $\mathsf{s}$ is a semilinear set,
    i.e., that there exists a finite set $Q$,
    numbers $\seqof{d_q}{q \in Q}$,
    vectors $\seqof{v_{i,q}}{0 \leq i \leq d_q, q \in Q}$ in $\Nat^{M^3}$,
    such that
    \begin{equation*}
        \mathsf{s}(\Sigma^*) =
        \bigcup_{q \in Q}
        \setof{
            v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
        }{
            (\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}
        } \quad .
    \end{equation*}

    We can therefore define a function $\mathsf{decode} \colon \Sigma^* \to
    \sum_{q \in Q} \Nat^{d_q}$ that maps to a word $w$ to \emph{some} $q \in Q$
    and parameters $(\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}$ such that
    $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$. By taming the
    semilinear set prior to this step, we can ensure that the function is in
    fact uniquely defined, this is done by computing an equivalent semilinear
    set that is \emph{unambiguous} and \emph{proper} \cite{CHHA16}.
    Similarly, let us define $\mathsf{encode} \colon \sum_{q \in Q} \Nat^{d_q}
    \to \Sigma^*$ that maps an element $q \in Q$ with parameters $(\lambda_1,
    \dots, \lambda_{d_q}) \in \Nat^{d_q}$ to \emph{some} word $w \in \Sigma^*$
    such that $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$.
    Notice that $\mathsf{decode} \circ \mathsf{encode} = \mathsf{id}$, by the
    unicity of the representation in $\sum_{q \in Q} \Nat^{d_q}$.
    We can define the map $\mathsf{repr} \colon \sum_{q \in Q} \Nat^{d_q} \to \Nat^{M^3}$
    as follows:
    \begin{equation*}
        \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \defined
        v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
        \quad .
    \end{equation*}
    Finally, let us define for $q \in Q$ the map $g_q \colon \Nat^{d_q} \to \Rel$ as the polynomial
    expression:
    \begin{equation*}
        g_q(\lambda_1, \dots, \lambda_{d_q}) \defined \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \quad .
    \end{equation*}

    We have proven that the following diagram commutes:
    \begin{center}
        \begin{tikzcd}
            \Sigma^* \arrow{rr}{f} 
                     \arrow{dr}{\mathsf{s}}
                     \arrow[bend right=20]{dd}[rotate=90,anchor=center,yshift=4pt]{\mathsf{decode}}
                     &        & \Rel \\
                     & \Nat^{M^3} \arrow{ur}[below]{g} & \\
            \sum_{q \in Q} \Nat^{d_q}
            \arrow[bend right=20]{uu}[right,rotate=90,anchor=center,yshift=-4pt]{\mathsf{encode}}
            \arrow{ur}[left]{\mathsf{repr}}
            \arrow[bend right=45]{uurr}[right]{\seqof{g_q}{q \in Q}}
                     & & \\
        \end{tikzcd}
    \end{center}

    The only missing part of this proof is the fact that the maps $\mathsf{encode}$ and
    $\mathsf{decode}$ can be computed as \kl{polyregular functions}. Notice
    that \emph{mutatis mutandis}, the same proof scheme applies to $f \in
    \ZSF[k]$, by assuming that the monoid is aperiodic, since the functions
    constructed (apart from $\mathsf{encode}$ and $\mathsf{decode}$) will be
    \kl{star-free}.
\end{proofof}

>>>>

%! TeX program = xelatex
%! lang = en-US

@@

@@

    We will first prove that \cref{d-t-transl:item} $\implies$
    \cref{d-t-correct:item}, for any choice of $K \in \Nat$. Indeed, let us
    consider some partial valuation $\nu \colon \vec{X} \topartial \Nat$. By
    assumption, $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$, and in
    particular conclude that \kl{maximal monomials} of
    $\translate{K}(\restr{P}{\nu})$ are exactly the \kl{maximal monomials} of
    $\restr{P}{\nu}$, and are therefore \kl{non-negative}.

    Conversely, the implication \cref{d-t-correct:item} $\implies$
    \cref{d-t-transl:item} is proven by induction on the number of
    indeterminates of $P$. If $P$ is a constant polynomial, then $P = n$ for
    some $n \in \Nat$, and we conclude that $\translate{0}(P) \in
    \Nat[\vec{X}]$. Otherwise, $P = P_1 + P_2$ where $P_1$ is the sum of the
    \kl{maximal monomials} of $P$. Using
    \cref{derivation-stabilises-correct:lem}, there exists a computable $K \in
    \Nat$ such that $Q \defined \Diff{K}{P_1} + \translate{K}(P_2)$ belongs to
    $\Nat[\vec{X}]$. Furthermore, for all partial functions $\nu \colon \vec{X}
    \topartial \set{0, \dots, K}$ fixing at least one indeterminate of $P$, the
    induction hypothesis yields a number $L_\nu \in \Nat$ such that
    $\translate{L_\nu}(\restr{P}{\nu}) \in \Nat[\vec{X}]$. Let us define $L_m$
    to be the maximum of $L_\nu$ where $\nu$ ranges over partial
    functions described above.

    Let us prove that for all partial functions $\nu \colon \vec{X} \topartial
    \Nat$, the translation $\translate{L_m + K}(\restr{P}{\nu})$ belongs to
    $\Nat[\vec{X}]$. Without loss of generality, $\nu$ fixes only
    indeterminates that appear in $P$. If $\nu$ fixes at least one
    indeterminate $X_i$ to a value $\nu(X_i) = k \leq K$, then we can write
    $\nu = \mu [X_i = k]$, where $\mu$ does not fix the value of $X_i$. By
    construction, we know that $\translate{L_i}(\restr{\restr{P}{X_i =
    k}}{\mu})$ belongs to $\Nat[\vec{X}]$, and therefore that $\translate{K +
    L_m}(\restr{P}{\nu})$ does so too. The only case left is when all the
    variables appearing in $\nu$ are greater than $K$.
    In that case, $\nu = \translate{K}(\mu)$ for some partial 
    valuation $\mu$.
    In this case,
    $\translate{K}(\restr{P}{\translate{K}(\mu)})
    = \restr{\translate{K}(P)}{\mu}$.
    Remark that by definition,
    $\translate{K}(P) = P_1 + Q$,
    hence that
    $\translate{K + L_m}(P) = \translate{L_m}(P_1) + \translate{L_m}(Q)$
    belongs to $\Nat[\vec{X}]$.
    Finally,
    we conclude that $\restr{\translate{K+L_m}(P)}{\mu}$
    belongs to $\Nat[\vec{X}]$
    because
    the latter is closed under partial applications.

@@

    Using this $\omega$,
    for any choice of
    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
    there exists two \kl{natural binomial functions} $F_r^+$ and $F_r^-$
    associated to $f_+$ and $f_-$,
    such that $X_1, \dots, X_p \geq 1$:
    \begin{equation*}
        f\left(
        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
        = F_r^+(X_1, \dots, X_p) - F_r^-(X_1, \dots, X_p) \quad .
    \end{equation*}
    Now, for large enough values of $x_1, \dots, x_p$,
    this means that 
    $f\left(
        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right)$
    coincides with an \kl{integer binomial polynomial},
    and we can therefore conclude selecting different values of $r$ and
    setting some subsets of variables to be assigned to the constant $0$.
\end{proofof}

@@

### 2024-08-15 parts/conclusion.tex
gigantic rewrite
enormous rewrite

Let us conclude with a more general remark on the status of commutative input
in the study of unary output polyregular functions. Notice that semantic
characterizations of subclasses of $\ZPoly$ (\kl{combinatorial}, \kl{ultimately
polynomial}) arise from pre-composition with a \kl{commutative}
\kl{star-free polyregular function}, as formally stated in
\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.

%Indeed, a suitable application of Parikh's Theorem \cite{PARI66}
%guarantees the existence of the functions $\mathsf{enc}$ and $\mathsf{dec}$,
%although they may fail to be  \kl{polyregular functions}.

\begin{lemma}
    \label{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$. Then,
    $f \in \ZPoly[k]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k\times l]$.
    \proofref{pre-compose-growth-commut:lemma}
\end{lemma}

\begin{lemma}
    \label{pre-compose-sf-commut:lemma}
    Let $f \in \ZPoly$. Then, $f \in \ZSF$,
    if and only if for every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
    \proofref{pre-compose-sf-commut:lemma}
\end{lemma}

Remark that if those two lemmas were to hold for \kl{$\Nat$-polyregular
functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
immediately, and $\NSF$ inside $\NPoly$ would immediately follow. On the one
hand, one can guess a candidate function and check for equivalence, on the
other hand, one can guess a \kl{commutative} \kl{star-free polyregular
function} and check membership (which is decidable thanks to this paper).

>>>>

Let us conclude with a more general remark on the status of commutative input
in the study of unary output polyregular functions. Throughout the paper, we
have tried to highlight that results on polyregular functions often have the
following form: a polyregular function $f \colon \Sigma^* \to \Gamma^*$
satisfies property $X$ if and only if for all \kl{commutative} \kl{star-free
polyregular function} $g \colon \Delta^* \to \Sigma^*$, $f \circ g$ satisfies
$X$. Let us formally restate this for two earlier results in the case of
\kl{$\Rel$-polyregular functions}
in \cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.

\begin{lemma}
    \proofref{pre-compose-growth-commut:lemma}
    \label{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$. Then,
    $f \in \ZPoly[k]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k\times l]$.
\end{lemma}

\begin{lemma}
    \proofref{pre-compose-sf-commut:lemma}
    \label{pre-compose-sf-commut:lemma}
    Let $f \in \ZPoly$. Then, $f \in \ZSF$,
    if and only if for every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
\end{lemma}

Remark that if those two lemmas were to hold for \kl{$\Nat$-polyregular
functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
immediately, and $\NSF$ inside $\NPoly$ would also immediately follow. On the
one hand, one can guess a candidate function and check for equivalence, on the
other hand, one can guess a \kl{commutative} \kl{star-free polyregular
function} and check membership (which is decidable thanks to this paper).
